---
title: "R Notebook"
output: html_notebook
---

```{r}
library(dplyr)
library(lubridate)
library(ggplot2)
```


```{r loading_data}
printing <- read.csv("~/remote/ocf_boc/data/printing_unan.csv", stringsAsFactors=FALSE, na.strings = "NULL")
sessions <- read.csv("~/remote/ocf_boc/data/sessions_uanon.csv", stringsAsFactors=FALSE, na.strings = "NULL")
staff_sessions <- read.csv("~/remote/ocf_boc/data/staff_session_duration_public.csv", stringsAsFactors=FALSE,na.strings = "NULL")
```


```{r}
get_time= function(string) {
  as.POSIXct(paste(substr(string, 0, 10), substr(string,12, 19)))
}

get_time_m = memoise::memoise(get_time)
```


```{r}
mar21_sessions = sessions %>% 
  filter(date(start) == "2018-03-21") %>%
  mutate(start = get_time_m(start)) %>% 
  mutate(end = get_time_m(end)) %>% 
  mutate(interval = lubridate::interval(start, end)) 

mar21_printing = printing %>% filter(date(time) == "2018-03-21") %>% 
  mutate(time = get_time_m(time)) 
```

```{r}
bold_agg_session = aggregate(data=mar21_sessions, interval~user, FUN=as.character)
bold_agg_printing = aggregate(data=mar21_printing, time~user, FUN=as.character)

bold_join =bold_agg_session %>% left_join(bold_agg_printing, by="user")
bold_join2 = (mar21_sessions %>% select(start, end, user, host))  %>% 
  left_join(mar21_printing %>% select(time, user), by="user") 
```

```{r}
mar21_sessions_merged = bold_join2 %>% 
  mutate(if_valid = (time>=start)&(time <= end)) 
#%>%
#  tidyr::replace_na(list(if_valid=TRUE))

temp = bold_join2 %>% 
  arrange(desc(time)) %>% 
  filter(!duplicated(cbind(start, end))) %>% 
  mutate(if_printing = (time>=start)&(time <= end)) %>% 
  tidyr::replace_na(list(if_printing=FALSE)) %>% 
  mutate(duration = as.numeric(round((end - start)/60, 1)))
  
```

```{r}
View(temp %>% select(if_printing, duration) %>% 
       group_by(if_printing) %>% 
       summarise(mean=mean(duration),
                 q10=quantile(duration, 0.1),
                 q25=quantile(duration, 0.25),
                 q50=quantile(duration, 0.5),
                 q75=quantile(duration, 0.75),
                 q90=quantile(duration, 0.9),
                 q95=quantile(duration, 0.95)))
#bold_join = (mar21_sessions %>% select(user, interval))%>% left_join(mar21_printing %>% select(time, user), by="user")
```

```{r}
ggplot(temp %>% filter(duration < 100)) +
  geom_histogram(aes(duration, fill=if_printing), alpha=0.8, position="identity", bins=50) + 
  scale_y_continuous(labels = scales::percent)
```

last semester
(date(start) >= "2017-08-24") & (date(start) <= "2017-12-15") 


```{r joining_table_on_global_scale}
printing <- read.csv("~/remote/ocf_boc/data/printing_unan.csv", stringsAsFactors=FALSE, na.strings = "NULL")
sessions <- read.csv("~/remote/ocf_boc/data/sessions_uanon.csv", stringsAsFactors=FALSE, na.strings = "NULL")
staff_sessions <- read.csv("~/remote/ocf_boc/data/staff_session_duration_public.csv", stringsAsFactors=FALSE,na.strings = "NULL")

sp18_sessions = sessions %>% 
  filter(end != "") %>%
  filter(date(start) >= "2018-01-16") %>%
# filter((date(start) >= "2017-08-24") & (date(start) <= "2017-12-15")) %>% 
  mutate(start = get_time_m(start)) %>% 
  mutate(end = get_time_m(end)) 

sp18_public_session = sp18_sessions %>% 
  filter(host != "blizzard.ocf.berkeley.edu") %>%
  filter(host != "eruption.ocf.berkeley.edu") %>%
  anti_join(staff_sessions, by="id")

sp18_printing = printing %>% 
  filter(date(time) >= "2018-01-16") %>% 
  mutate(time = get_time_m(time)) 
#  filter((date(time) >= "2017-08-24") & (date(time) <= "2017-12-15")) 
  

#bold_agg_session = aggregate(data=sp18_sessions, interval~user, FUN=as.character)
#bold_agg_printing = aggregate(data=sp18_printing, time~user, FUN=as.character)

#bold_join =bold_agg_session %>% left_join(bold_agg_printing, by="user")
bold_join2 = (sp18_public_session %>% select(start, end, user, host))  %>% 
  left_join(sp18_printing %>% select(time, user), by="user") 

temp = bold_join2 %>% 
  arrange(desc(time)) %>% 
  filter(!duplicated(cbind(start, end))) %>% 
  mutate(if_printing = (time>=start)&(time <= end)) %>% 
  tidyr::replace_na(list(if_printing=FALSE)) %>% 
  mutate(duration = as.numeric(round((end - start)/60, 1))) %>% 
  filter(duration > 0)

ggplot(temp %>% filter((hour(start) >= 10) & (hour(end) < 18))) +
  geom_histogram(aes(duration, fill=if_printing), alpha=0.8, position="identity", bins=50) +
  facet_grid(. ~ weekdays(start))  + xlim(0, 100)

ggplot(temp %>% filter((hour(start) >= 9) & (hour(end) < 18))) +
geom_density(aes(duration, fill=if_printing), alpha=0.3, position="identity") +
facet_grid(. ~ weekdays(start)) + xlim(0, 100)

ggplot(temp %>% filter((hour(start) >= 9) & (hour(end) < 18))) +
stat_ecdf(aes(duration, color=if_printing), position="identity") +
facet_grid(. ~ weekdays(start)) + xlim(0, 100) + ylim(0.5, 1)
```

```{r}
temp_table = temp %>% filter((hour(start) >= 10) & (hour(end) <= 18)) %>% select(if_printing, duration, start) %>% 
       group_by(if_printing, weekdays(start)) %>% 
       summarise(mean=mean(duration),
                 q10=quantile(duration, 0.1),
                 q25=quantile(duration, 0.25),
                 q50=quantile(duration, 0.5),
                 q55=quantile(duration, 0.55),
                 q60=quantile(duration, 0.6),
                 q65=quantile(duration, 0.65),
                 q70=quantile(duration, 0.70),
                 q75=quantile(duration, 0.75),
                 q90=quantile(duration, 0.9),
                 q95=quantile(duration, 0.95), n())
```


```{r}
get_quantiles = function(x){
  percentiles = seq(0.1, 1-0.05, 0.05)
  paste(round(quantile(x, percentiles),1), collapse = "||")
}

get_quantiles(1:100)
```

```{r}
temp %>% filter((hour(start) >= 9) & (hour(end) < 18)) %>%
  select(if_printing, duration, start) %>% 
  group_by(if_printing, weekdays(start)) %>% 
  summarise(x = get_quantiles(duration))
```

```{r}
print_by_dates =temp %>% filter((hour(start) >= 9) & (hour(end) < 18))  %>% 
  group_by(date(start), weekdays(start)) %>% 
  count(if_printing) %>% 
  tidyr::spread(if_printing, n) %>% 
  mutate(not_print = `FALSE`/(`FALSE` + `TRUE`)) %>% 
  mutate(print = `TRUE`/(`FALSE` + `TRUE`))


print_by_weeks =temp %>% filter((hour(start) >= 9) & (hour(end) < 18))  %>% 
  group_by(week(start)) %>% 
  count(if_printing) %>% 
  tidyr::spread(if_printing, n) %>% 
  mutate(not_print = `FALSE`/(`FALSE` + `TRUE`)) %>% 
  mutate(print = `TRUE`/(`FALSE` + `TRUE`))
```

```{r}
ggplot(print_by_dates) +
  geom_line(aes(x=`date(start)`,y=print), colour="red") +
  geom_line(aes(x=`date(start)`,y=1-print), colour="blue") +
  ylim(0, 1)

ggplot(print_by_weeks) +
  geom_line(aes(x=`week(start)`+30,y=print), colour="red") +
  geom_line(aes(x=`week(start)`+30,y=1-print), colour="blue") +
  ylim(0, 1)
```

